Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-02 text/manuscript-full-KS.qmd
methods revisions

**RQ1**. RQ1 primarily utilizes Census Data while RQ2 primarily utilizes data about student list purchases. Research question 1 is about socioeconomic and racial variation between Geomarkets in metropolitan areas and how such variation changes over time. To answer this question, we use census tract data from the U.S. Census Bureau. Utilizing a small geographic area is important for implementing a spatial merge between Census data and Geomarket shape files (described below). Census tract is the smallest geographic area for which measures of both race/ethnicity and socioeconomic characteristics are available. Census tracts are statistical subdivisions of a county designed to capture relatively homogeneous demographic communities that are generally considered to be more accurate for population analyses than zip codes, which are spatial boundaries drawn to demarcate route efficiency for postal services \[[CITE](https://www.census.gov/newsroom/blogs/random-samplings/2014/07/understanding-geographic-relationships-counties-places-tracts-and-more.html)\] .

@@

We assigned each census tract to College Board Geomarkets by implementing a partial spatial merge. Spatial manipulations utilized functions from the *sf* R package. Data to create the shapefiles for the College Board Geomarkets were obtained from a 2012 R-bloggers post written by a data scientist working in the enrollment management space.[^17] Essentially, the data file had one observation per zip code and a column that assigned each zip code to one Geomarket. Using these zip code-level data, we utilized the `aggregate()` function to create shapefiles for each Geomarket.

[^17]: The author of the post reported that he found the zip codes associated with Geomarkets from a Google search.

Next, we utilized the `st_intersection()` function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. We split census-tract shapefiles that intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets) into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people.

@@

**Research design**. We utilize a multiple, quantitative case study design. Metropolitan areas are cases. Case studies deepen understanding of phenomena by comparing and contrasting across different settings \[CITE YIN 2001\]. Our research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to generalize Geomarket variation nationally, but rather to explore how differences across Geomarket landscapes shape prospect pools based on the geographical model of metropolitan areas and their demographic characteristics \[MAKE THIS MORE ABOUT EXPLORING THE MODEL. THIS SENTENCE SOUNDS FANCY BUT DOESN'T REALLY SAY ANYTHING\]. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomenon in rich and nuanced contexts. We, therefore, select a small number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics \[TIE IN THEORETICAL FRAMEWORK HERE A BIT MORE\]. Online Appendix XXX presents results for a larger number of metropolitan areas.

@@

@fig-chicago-rq2-highSAT-combo shows the magnitude of Geomarket contribution disparities increases at the highest SAT score range (1310-1600), partly driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories. -->

```
@fig-chicago-rq2-midSAT-race shows which Geomarkets send the bulk of students for a given race/ethnicity. The City of Chicago accounts for 17% of all prospects but 48% of Black students. Similarly, South & Southwest Suburbs accounts for 20% of all prospects but 30% of all Black prospects. The upshot is that if the student list purchases had excluded these two Geomarkets, 78% of the nearly 1,000 Black students currently included in these purchases would be excluded. On the other hand, all other Geomarkets contributed substantially smaller shares of Black prospects. For example, the Northwest Suburbs accounts for about 13% of all prospects but only about 3% of Black prospects. The City of Chicago Geomarket also contributed larger shares of the nearly 3,000 Hispanic students (39% versus 17% for all prospects), while all other Geomarkets contributed comparable or relatively smaller proportions of Hispanic prospects than their overall representation among all included prospects.

>>>>

**RQ1**. RQ1 primarily utilizes Census Data while RQ2 primarily utilizes data about student list purchases. Research question 1 is about socioeconomic and racial variation between Geomarkets in metropolitan areas and how such variation changes over time. To answer this question, we use census tract data from the U.S. Census Bureau. Utilizing a small geographic area is important for implementing a spatial merge between Census data and Geomarket shape files (described below). Census tracts are the smallest geographic area for which measures of both race/ethnicity and socioeconomic characteristics are available. Defined as statistical subdivisions of counties, census tracts boundaries are reassessed every decennial census to reflect relatively homogeneous demographic communities, typically ranging in population size from 2,500 and 8,000 residents [@Bureau2014; @Bureau1994]. Compared to zip codes, which are fixed spatial units designed primarily to demarcate route efficiency for postal services, census tracts are considered more reliable for population-based analyses.

@@

We assigned each census tract to a College Board Geomarket, which we constructed through manual digitization of maps published by the College Board in 2023 marketing materials promoting their Enrollment Planning Service (EPS) software products. Specifically, we georeferenced the College Board’s Geomarket maps by overlaying PDF images onto spatial coordinates aligned with census tract boundaries. We then extracted vector data by manually tracing the Geomarket boundaries and generating individual shapefiles for each Geomarket.

This approach to constructing Geomarkets aligned precisely with 2020 Decennial Census tract boundaries; that is, each census tract corresponded uniquely to a Geomarket generated from the College Board images. However, this process was less precise when applied to the 1980 and 2000 Decennial Census tract boundaries, likely a factor of census tract delineations shifting over time based on demographic changes and our use of map images from 2023. For instance, census tracts that experience population growth exceeding the 8,000 resident threshold can be split into two or more smaller census tracts to maintain consistent population ranges. Conversely, two or more adjacent census tracts experiencing population declines can sometimes be merged to form a single census tract whose population falls within the 2,500 to 8,000 resident range.

Some census tract boundaries from the 1980 and 2000 Decennial Censuses intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets), resulting in spatial misalignment. For tracts divided across multiple Geomarkets, we assigned each tract to the Geomarket that encompassed the majority of its spatial area. This method introduced slight differences to Geomarket shapes across decennial censuses. However, to assess the robustness of our approach to spatial misalignments, we tested alternative methods for resolving ambiguous tract assignments to Geomarkets and found overall results remained relatively consistent across approaches.[^17]

[^17]: An alternative approach to dealing with spatial misalignments included creating shapefiles for Geomarkets from a 2012 R-bloggers post written by a data scientist working in the enrollment management space. Essentially, the data file was a Geomarket-zipcode crosswalk that held one observation per zip code and a column that assigned each zip code to one Geomarket. Using these zip code-level data, we utilized the aggregate() function to create shapefiles for each Geomarket. Next, we utilized the st_intersection() function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. This spatial merge kept Geomarket boundaries consistent across all three Decennial Census and split census-tract shapefiles that intersected multiple Geomarkets into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people. Even under this most conservative approach to addressing spatial misalignment, overall patterns in the findings remained consistent. However, we opted to georeference the manually digitized maps directly sourced from the College Board, as the original source of the Geomarket-zipcode crosswalk cited in the the R-bloggers post could not be verified.

```{=html}
<!--#We assigned each census tract to College Board Geomarkets by implementing spatial merges. Spatial manipulations utilized functions from the sf R package. Data to create the shapefiles for the College Board Geomarkets were obtained from a 2012 R-bloggers post written by a data scientist working in the enrollment management space. Essentially, the data file had one observation per zip code and a column that assigned each zip code to one Geomarket. Using these zip code-level data, we utilized the aggregate() function to create shapefiles for each Geomarket.

Next, we utilized the st_intersection() function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. We split census-tract shapefiles that intersected multiple Geomarkets (i.e., tracts whose boundaries spanned two or more Geomarkets) into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people.
--> 
```

@@

**Research design**. We utilize a multiple, quantitative case study design. Metropolitan areas are cases. Case studies deepen understanding of phenomena by comparing and contrasting across different settings [@RN4011]. Our research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to generalize Geomarket variation nationally, but rather to explore how differences across Geomarket landscapes shape prospect pools based on the geographical model of metropolitan areas and their demographic characteristics \[MAKE THIS MORE ABOUT EXPLORING THE MODEL. THIS SENTENCE SOUNDS FANCY BUT DOESN'T REALLY SAY ANYTHING\]. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomenon in rich and nuanced contexts. We, therefore, select a small number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics \[TIE IN THEORETICAL FRAMEWORK HERE A BIT MORE\]. Online Appendix XXX presents results for a larger number of metropolitan areas.

@@

@fig-chicago-rq2-highSAT-combo shows the magnitude of Geomarket contribution disparities increases at the highest SAT score range (1310-1600), partly driven by the relatively fewer number of first-generation college students included within these orders. Again, the City of Chicago Geomarket disproportionately contributes a disproportionate share of first-generation Asian college students (43% across both parental education categories) and first-generation Hispanic college students (52% across both parental education categories) in comparison to the overall distribution of first-generation college students within each racial/ethnic category (13% for Asian and 40% for Hispanic, respectively). Conversely, nearly all other Geomarkets contribute a disproportionately larger share of non-first generation college students across all racial/ethnic categories. -->
```
@fig-chicago-rq2-midSAT-race shows which Geomarkets send the bulk of students for a given race/ethnicity. The City of Chicago accounts for 17% of all prospects but 48% of Black students. Similarly, South & Southwest Suburbs accounts for 20% of all prospects but 30% of all Black prospects. The upshot is that if the student list purchases had excluded these two Geomarkets, 78% of the nearly 1,000 Black students currently included in these purchases would be excluded. On the other hand, all other Geomarkets contributed substantially smaller shares of Black prospects. For example, the Northwest Suburbs accounts for about 13% of all prospects but only about 3% of Black prospects. The City of Chicago Geomarket also contributed larger shares of the nearly 3,000 Hispanic students (39% versus 17% for all prospects), while all other Geomarkets contributed comparable or relatively smaller proportions of Hispanic prospects than their overall representation among all included prospects.
